---
title: "Introduction to FedGEE"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Introduction to FedGEE}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

## Overview

**FedGEE** fits a generalized estimating equation across sites that never
pool patient data. Each site sends only $p \times p$ summaries: its bread
$B_i$ and its score $S_i$. There are two ways to combine them:

- `fedgee()`: a central server sums the summaries. The estimate equals pooled
  GEE.
- `decentralized_fedgee()`: there is no server. Sites average with their
  neighbours over a network (gossip). With enough rounds it reproduces
  `fedgee()`.

With few sites the usual sandwich is too small. FedGEE corrects it in score
space, using only the site summaries.

```{r setup}
library(FedGEE)
data(ChickWeight)
cw <- as.data.frame(ChickWeight)
cw$site <- as.integer(cw$Chick) %% 12
data_list <- split(cw, cw$site)
```

## Centralized fit

One call computes every correction. The printout shows the default:
Kauermann–Carroll (KC) with Bell–McCaffrey degrees of freedom.

```{r fit}
fit <- fedgee(
  data_list, weight ~ Time + Diet,
  family_obj = gaussian(), id_col = "Chick", verbose = FALSE
)
fit
```

You can show any other variant without refitting:

```{r variants}
summary(fit, correction = "MD", df = "K-1")
sapply(fit$variances$site, function(v) round(v$se, 3))
```

The point estimate and the uncorrected sandwich match pooled GEE clustered
by site:

```{r pooled}
pooled <- geepack::geeglm(weight ~ Time + Diet,
  data = cw[order(cw$site), ], id = site,
  family = gaussian(), corstr = "independence"
)
all.equal(unname(coef(fit)), unname(coef(pooled)))
all.equal(unname(vcov(fit, "none")), unname(vcov(pooled)))
```

### Which correction?

- **KC + `df = "bm"`** (the default). KC is unbiased under the working model.
  The Bell–McCaffrey df adapts to unequal site sizes. In simulation this pair
  gave close to nominal coverage in every setting.
- **MD** is conservative. **FG** is a diagonal alternative. It rarely
  differs from KC unless one site has very high leverage (see
  `fit$diagnostics$max_leverage`).
- The site-level sandwich has rank $\min(p, K-1)$. Keep $p$ well below the
  number of sites. `fedgee()` warns when the sandwich is singular.

## Decentralized fit

`build_weight_matrix()` builds the gossip matrix and reports its mixing rate
$\rho$. Disagreement between sites shrinks like $\rho^L$ after $L$ rounds.

```{r network}
sapply(c("complete", "hub", "ring"), function(s) build_weight_matrix(12, s)$rho)
```

On a ring, a few rounds per iteration leave the sites disagreeing. More
rounds recover the centralized fit:

```{r dec}
few <- decentralized_fedgee(data_list, weight ~ Time + Diet,
  family_obj = gaussian(), id_col = "Chick", structure = "ring",
  sandwich_level = "site", correction = "KC",
  L_beta = 5, L_S = 5, L_B = 5, verbose = FALSE
)
many <- decentralized_fedgee(data_list, weight ~ Time + Diet,
  family_obj = gaussian(), id_col = "Chick", structure = "ring",
  sandwich_level = "site", correction = "KC",
  L_beta = 150, L_S = 150, L_B = 150, tol = 1e-6, verbose = FALSE
)
rbind(
  centralized = fit$se,
  ring_L5 = few$se,
  ring_L150 = many$se
)
```
