---
title: "Simulation Study"
format:
  html:
    theme: cosmo
    toc: true
    number-sections: false
    grid:
      body-width: 1200px
---

<style>
  p {
    text-align: justify;
  }
</style>

```{r}
#| label: setup
#| include: false

library(dplyr)
library(tidyr)
library(ggplot2)
library(patchwork)
library(scales)
library(knitr)

# ── Load & combine all result files ──────────────────────────────────────────
res_dir <- "~/2026_fedGee/simulations/sim_results"

files <- sprintf(file.path(res_dir, "results_%04d.rds"), 1:300)
exist <- file.exists(files)

all_res <- bind_rows(lapply(files[exist], readRDS))

# Drop any ERROR rows
all_res <- all_res |> filter(method != "ERROR")
```

## Simulation Study

To evaluate the statistical properties of the Federated Generalized Estimating Equation (FedGEE) algorithm, we conducted a comprehensive simulation study. The primary objectives were to demonstrate the mathematical equivalence between FedGEE and centralized pooled GEE, evaluate its robustness against naive meta-analytical approaches under complex clustering, and assess the performance of its variance estimators across varying network sizes and correlation structures.

### Data-Generating Process

We simulated a realistic hierarchical data structure reflecting multi-center clinical data: longitudinal visits nested within patients, who are in turn nested within hospitals. Let $Y_{ijk}$ denote the binary outcome for visit $k$ ($k = 1, \dots, n_{ij}$) of patient $j$ ($j = 1, \dots, N_i$) at hospital $i$ ($i = 1, \dots, K$). The true marginal probability of the outcome was specified via the logistic model:

$$
\text{logit}(p_{ijk}) = \beta_0 + \beta_X X_{ijk} + \beta_{\text{teach}} \text{Teaching}_i
$$

where $X_{ijk} \sim \mathcal{N}(0, 1)$ is a continuous, observation-level covariate, and $\text{Teaching}_i \in \{0, 1\}$ is a site-level (hospital-constant) covariate denoting whether the $i$-th hospital is a teaching facility (1) or a non-teaching facility (0). To induce the hierarchical correlation structure while strictly preserving the specified marginal probabilities, we utilized a latent Gaussian thresholding approach. A latent variable $Z_{ijk}$ was generated for each observation as:

$$
Z_{ijk} = \sqrt{\rho_H} U_i + \sqrt{\rho_P - \rho_H} W_{ij} + \sqrt{1 - \rho_P} \epsilon_{ijk}
$$

where $U_i, W_{ij}, \epsilon_{ijk} \sim \mathcal{N}(0, 1)$ are mutually independent standard normal variables representing hospital-level, patient-level, and observation-level variations, respectively. The parameters $\rho_H$ and $\rho_P$ control the intra-class correlation at the hospital and patient levels (with $\rho_P \ge \rho_H$ to ensure valid, non-negative variance components for the nested patient effects). The binary outcome was then generated by thresholding the latent variable: $Y_{ijk} = I(Z_{ijk} < \Phi^{-1}(p_{ijk}))$, where $\Phi^{-1}$ is the standard normal quantile function.

Simulations were run across a comprehensive parameter grid. The number of hospitals ($K$) ranged from 5 to 100. Each hospital treated a random number of patients uniformly distributed between 10 and 30, and each patient had a random number of visits uniformly distributed between 2 and 6. The true covariate effect was fixed at $\beta_X = 1$. To evaluate performance under varying baseline risks, the intercept was set to $\beta_0 = -1$ (common outcome) and $\beta_0 = -3$ (rare outcome). The between-hospital correlation ($\rho_H$) varied from 0 to 0.20, and the total patient-level correlation ($\rho_P$) varied from 0.225 to 0.40. Each unique parameter combination was replicated 200 times.

### Evaluated Methods

For each simulated dataset, we evaluated the treatment effect ($\beta_X$) using six distinct analytical approaches:

1. **Pooled GEE**: A standard, centralized GEE fitted on the fully aggregated dataset, utilizing an independence working correlation matrix and hospital-level clustering for the robust sandwich estimator.
2. **FedGEE (Site Sandwich)**: Our proposed federated algorithm utilizing an independence working correlation matrix and the aggregated hospital-level outer product to construct the sandwich variance.
3. **FedGEE (Site Sandwich + MD)**: The federated algorithm incorporating the Mancl-DeRouen (MD) small-sample leverage correction applied to the final global update, with inference drawn from a $t$-distribution ($df = K - p$).
4. **FedGEE (Patient Sandwich)**: The federated algorithm constructing the sandwich variance strictly from the sum of patient-level outer products, assuming independence across patients.
5. **Meta-Analytic GLM**: A traditional distributed approach where standard logistic regression models are fitted locally at each hospital, and the resulting coefficients are aggregated centrally using inverse-variance weighting. 
6. **GLMM**: A centralized generalized linear mixed-effects model incorporating random intercepts for both hospitals and patients (for estimand comparison purposes).

### Performance Metrics

Model performance across the 200 replications for each scenario was evaluated using the following metrics:

* **Bias**: The mean difference between the estimated coefficient and the true marginal parameter ($\bar{\hat{\beta}} - \beta_{\text{true}}$).
* **Empirical Standard Deviation (SD)**: The standard deviation of the parameter estimates across replications, representing the true sampling variability.
* **Mean Estimated Standard Error (SE)**: The average of the model-estimated standard errors.
* **SE/SD Ratio**: A measure of variance estimation accuracy, where a ratio of 1 indicates perfect estimation, $<1$ indicates anti-conservative standard errors, and $>1$ indicates conservative standard errors.
* **Coverage Probability**: The proportion of replications in which the 95% confidence interval contained the true parameter value (utilizing $z$-based limits for uncorrected models and $t$-based limits for the MD-corrected model).

### Simulation Deliverables

The results of the simulation study are organized to address the accuracy of point estimation across network conditions (@fig-sim-bias) and the robustness of the variance estimators across network sizes (@tbl-sim-coverage).

<br>

```{r}
#| label: fig-sim-bias
#| fig-cap: "**Point Estimation and Bias in Federated vs. Traditional Networks.** All panels use $K = 25$ and $\\beta_0 = -1$ (common outcome) unless otherwise noted. **(A)** Mean absolute bias heatmap of FedGEE vs. Meta-Analytic GLM across the full $\\rho_H \\times \\rho_P$ grid, illustrating the degradation of traditional meta-analysis as between-hospital correlation grows. **(B)** Mean absolute bias under a rare outcome ($\\beta_0 = -3$), demonstrating the robustness of the federated Newton-Raphson update when local site data is sparse. **(C)** Histogram of the site-level covariate estimates ($\\hat{\\beta}_{\\text{teach}}$) generated by FedGEE, showing the algorithm's ability to accurately recover site-constant effects strictly from between-site variation. **(D)** Scatterplot establishing mathematical parity: FedGEE coefficient estimates plotted directly against Centralized Pooled GEE estimates. The maximum absolute difference across all replications in this scenario is annotated."
#| fig-width: 14
#| fig-height: 11
#| echo: false
#| warning: false
#| message: false

# ── Shared heatmap builder for Panels A & B ──────────────────────────────────
build_bias_heatmap <- function(data, b0, panel_label, panel_subtitle) {
  
  hm_dat <- data |>
    filter(beta_0 == b0, K == 25,
           include_teaching == FALSE,
           method %in% c("fed_site", "meta_glm"),
           !is.na(bias)) |>
    group_by(rho_H, rho_P, method) |>
    summarise(mean_abs_bias = mean(abs(bias)), .groups = "drop") |>
    mutate(
      method = recode(method,
                      fed_site = "FedGEE",
                      meta_glm = "Meta-Analytic GEE"),
      rho_H = factor(rho_H),
      rho_P = factor(rho_P)
    )
  
  ggplot(hm_dat, aes(x = rho_P, y = rho_H, fill = mean_abs_bias)) +
    geom_tile(colour = "white", linewidth = 0.4) +
    geom_text(aes(label = sprintf("%.3f", mean_abs_bias)),
              size = 2.8, colour = "black") +
    facet_wrap(~ method) +
    scale_fill_distiller(palette = "YlOrRd", direction = 1,
                         name = "Mean |Bias|") +
    labs(x = expression(rho[P]),
         y = expression(rho[H]),
         title = bquote(bold(.(panel_label)) ~ .(panel_subtitle))) +
    theme_bw(base_size = 12) +
    theme(plot.title       = element_text(size = 12),
          legend.position  = "bottom",
          legend.key.width = unit(1.2, "cm"),
          panel.grid       = element_blank(),
          strip.background = element_rect(fill = "grey92"),
          strip.text       = element_text(face = "bold"))
}

# ── Panel A: Common outcome heatmap ──────────────────────────────────────────
panel_a <- build_bias_heatmap(all_res, b0 = -1,
                               panel_label = "A",
                               panel_subtitle = "  Common outcome (K = 25)")

# ── Panel B: Rare outcome heatmap ────────────────────────────────────────────
panel_b <- build_bias_heatmap(all_res, b0 = -3,
                               panel_label = "B",
                               panel_subtitle = "  Rare outcome (K = 25)")

# ── Panel C: Histogram of beta_teach estimates (K = 25, β₀ = −1) ─────────────
teach_hat <- all_res |>
  filter(include_teaching == TRUE,
         method == "fed_site",
         K == 25, beta_0 == -1,
         !is.na(beta_teach_hat))

true_teach <- 0.5

panel_c <- ggplot(teach_hat, aes(x = beta_teach_hat)) +
  geom_histogram(aes(y = after_stat(density)),
                 bins = 50, fill = "#2166AC", colour = "white", alpha = 0.8) +
  geom_vline(xintercept = true_teach, linetype = "dashed",
             colour = "#B2182B", linewidth = 0.9) +
  annotate("text", x = true_teach, y = Inf,
           label = paste0("True = ", true_teach),
           hjust = -0.15, vjust = 1.5, colour = "#B2182B", size = 3.5) +
  labs(x = expression(hat(beta)[teach]),
       y = "Density",
       title = expression(bold("C") ~ "  Recovery of site-level covariate")) +
  theme_bw(base_size = 12) +
  theme(plot.title = element_text(size = 12))

# ── Panel D: FedGEE vs Pooled GEE parity scatterplot (K = 25, β₀ = −1) ──────
parity <- all_res |>
  filter(method %in% c("fed_site", "pooled_gee"),
         K == 25, beta_0 == -1,
         !is.na(beta_hat)) |>
  select(rho_H, rho_P, K, beta_0, include_teaching, rep, method, beta_hat) |>
  pivot_wider(names_from = method, values_from = beta_hat) |>
  filter(!is.na(fed_site), !is.na(pooled_gee))

max_diff <- max(abs(parity$fed_site - parity$pooled_gee))

panel_d <- ggplot(parity, aes(x = pooled_gee, y = fed_site)) +
  geom_abline(intercept = 0, slope = 1, linetype = "dashed",
              colour = "#B2182B", linewidth = 0.6) +
  geom_point(alpha = 0.15, size = 0.4, colour = "#2166AC") +
  annotate("text", x = -Inf, y = Inf,
           label = sprintf("Max |diff| = %.1e", max_diff),
           hjust = -0.08, vjust = 1.5, size = 3.2, colour = "grey30") +
  labs(x = expression("Pooled GEE" ~ hat(beta)[X]),
       y = expression("FedGEE" ~ hat(beta)[X]),
       title = expression(bold("D") ~ "  Mathematical parity")) +
  theme_bw(base_size = 12) +
  theme(plot.title = element_text(size = 12))

# ── Compose (equal panel widths & heights) ───────────────────────────────────
(panel_a | panel_b) / (panel_c | panel_d) +
  plot_layout(heights = c(1, 1), widths = c(1, 1)) &
  theme(legend.position = "bottom")
```

<br>

```{r}
#| label: tbl-sim-coverage
#| tbl-cap: "Performance of Variance Estimators by Number of Network Sites ($K$). Coverage probabilities and Standard Error to Empirical Standard Deviation (SE/SD) ratios are shown for varying sandwich architectures and small-sample corrections within the FedGEE framework. Target coverage is 0.950; target SE/SD ratio is 1.00. *Note*: Values reflect simulations where $\\rho_H > 0$. Patient-level sandwich assumes independence between patients. The SE/SD ratio is identical for the site-level $z$ and $t$ rows (same standard error estimate; only the critical value differs) and is reported once. MD = Mancl-DeRouen correction."
#| echo: false
#| warning: false
#| message: false

# ── Patient-level sandwich (z-based) ────────────────────────────────────────
tbl_patient <- all_res |>
  filter(rho_H > 0, method == "fed_patient",
         !is.na(beta_hat), !is.na(se_hat), !is.na(covers_z)) |>
  group_by(K) |>
  summarise(
    cov_patient  = mean(covers_z),
    sesd_patient = mean(se_hat) / sd(beta_hat),
    .groups = "drop"
  )

# ── Site-level sandwich, z-based (fed_site) ──────────────────────────────────
tbl_site_z <- all_res |>
  filter(rho_H > 0, method == "fed_site",
         !is.na(beta_hat), !is.na(se_hat), !is.na(covers_z)) |>
  group_by(K) |>
  summarise(
    cov_site_z  = mean(covers_z),
    sesd_site   = mean(se_hat) / sd(beta_hat),
    .groups = "drop"
  )

# ── Site-level sandwich, t-based (fed_site) ──────────────────────────────────
tbl_site_t <- all_res |>
  filter(rho_H > 0, method == "fed_site",
         !is.na(beta_hat), !is.na(se_hat), !is.na(covers_t)) |>
  group_by(K) |>
  summarise(
    cov_site_t = mean(covers_t),
    .groups = "drop"
  )

# ── Site-level + MD correction, t-based (fed_site_md) ────────────────────────
tbl_md <- all_res |>
  filter(rho_H > 0, method == "fed_site_md",
         !is.na(beta_hat), !is.na(se_hat), !is.na(covers_t)) |>
  group_by(K) |>
  summarise(
    cov_md  = mean(covers_t),
    sesd_md = mean(se_hat) / sd(beta_hat),
    .groups = "drop"
  )

# ── Combine ──────────────────────────────────────────────────────────────────
tbl_final <- tbl_patient |>
  left_join(tbl_site_z, by = "K") |>
  left_join(tbl_site_t, by = "K") |>
  left_join(tbl_md,     by = "K") |>
  arrange(K)

# ── Format for display ──────────────────────────────────────────────────────
tbl_display <- tbl_final |>
  transmute(
    `$K$`                                        = K,
    `Patient Sandwich<br>Coverage`               = sprintf("%.3f", cov_patient),
    `Patient Sandwich<br>SE/SD`                  = sprintf("%.2f", sesd_patient),
    `Site Sandwich ($z$)<br>Coverage`             = sprintf("%.3f", cov_site_z),
    `Site Sandwich<br>SE/SD`                      = sprintf("%.2f", sesd_site),
    `Site Sandwich ($t$)<br>Coverage`             = sprintf("%.3f", cov_site_t),
    `Site + MD ($t$)<br>Coverage`                 = sprintf("%.3f", cov_md),
    `Site + MD ($t$)<br>SE/SD`                    = sprintf("%.2f", sesd_md)
  )

kable(tbl_display, align = c("l", rep("c", 7)), escape = FALSE)
```

### Simulation Results and Methodological Justification

The simulation study provides compelling justification for deploying the FedGEE architecture over traditional distributed analytical methods, specifically highlighting the necessity of the global Newton-Raphson update and the small-sample variance corrections. 

**The Failure of Traditional Meta-Analysis**
As illustrated in @fig-sim-bias, traditional meta-analytic approaches—where models are fitted locally and aggregated via inverse-variance weighting—degrade significantly in the presence of hierarchical clinical data. Under a common outcome ($\beta_0 = -1$), the meta-analytic GLM exhibited increasing bias as the between-hospital correlation ($\rho_H$) grew. More critically, when evaluating a rare outcome ($\beta_0 = -3$), the local meta-analytic models frequently failed to converge due to sparse local data, rendering the approach fundamentally unviable for studying rare clinical events across a federated network. In stark contrast, FedGEE maintained exact mathematical parity with the centralized pooled GEE. As shown in Panel D for the reference scenario ($K = 25$, $\beta_0 = -1$), the maximum absolute difference between federated and centralized estimates was negligible (on the order of machine precision); this parity held identically across all simulated scenarios. By aggregating the Score and Bread matrices centrally rather than the coefficients themselves, FedGEE perfectly solved the global estimating equations, remaining strictly unbiased regardless of the underlying correlation structure or baseline outcome prevalence.

**Recovery of Hospital-Level Interventions**
A critical limitation of perfectly siloed local models is their mathematical inability to estimate the effects of site-level (hospital-constant) covariates, such as a facility's teaching status or complexity. Because these variables have zero variance within a single site, local models drop them to avoid singularity. As demonstrated in @fig-sim-bias (Panel C), FedGEE seamlessly resolves this limitation. By initializing site-constant covariates at zero locally and estimating their effects purely through between-site variation during the server-level update, FedGEE successfully and unbiasedly recovered the true effect of the site-level covariate ($\beta_{\text{teach}}$). This capability is absolutely essential for health systems evaluating system-level structural interventions across multiple facilities.

**Ensuring Valid Inference in Small Networks**
Accurate point estimation must be accompanied by robust variance estimation to prevent inflated Type I errors. As detailed in @tbl-sim-coverage, the choice of the "meat" in the sandwich variance estimator is paramount. The patient-level sandwich estimator—which assumes independence among patients—resulted in severe under-coverage (confidence intervals that were too narrow) whenever hospital-level clustering ($\rho_H > 0$) was present. 

While upgrading to the site-level sandwich estimator correctly accounted for within-hospital correlation, the standard asymptotic $z$-based limits still undercover when the number of network sites ($K$) is small to moderate (e.g., $K < 50$), a common reality in federated health networks. The application of the Mancl-DeRouen (MD) leverage correction, combined with inference drawn from a $t$-distribution, successfully rescued the coverage probability. The MD-corrected FedGEE maintained the nominal 95% coverage rate and an ideal SE/SD ratio of roughly 1.00 across all network sizes, proving its reliability even in highly restricted federated topologies (e.g., $K = 10$).

**Conclusion**
Together, these results confirm that analyzing clustered clinical data without centralizing it requires more than simple model averaging. The FedGEE algorithm—configured with a site-level sandwich and Mancl-DeRouen correction—is strictly required to achieve the exact statistical accuracy of a pooled database, rigorously adjust for differing local case mixes, and evaluate hospital-level characteristics, all while maintaining a zero-trust, privacy-preserving data architecture.
