---
title: "Federated GEE for Clustered Binary Outcomes: Simulation Studies"
subtitle: "AUD Medication Prescribing Across VA Hospitals"
author: ""
format:
  html:
    page-layout: full
execute:
  echo: true
  warning: false
  message: false
  cache: true
---

```{r}
#| label: setup
#| code-fold: true
#| code-summary: "Load libraries"

library(lme4)
library(geepack)
library(dplyr)
library(tidyr)
library(ggplot2)
library(mvtnorm)
library(patchwork)

set.seed(2026)
```

# Data Generation

The data-generating process must produce correlated binary outcomes while preserving exact marginal probabilities. We use a nested latent normal (copula) construction that induces correlation at both the hospital and patient levels without distorting the marginal model.

## The Nested Copula Construction

For patient $i$ at hospital $h$, visit $j$, the latent variable is:

$$
Z_{hij} = \sqrt{\rho_H}\; U_h + \sqrt{\rho_P - \rho_H}\; W_{hi} + \sqrt{1 - \rho_P}\; \epsilon_{hij}
$$

where $U_h \sim N(0,1)$ is shared across hospital $h$, $W_{hi} \sim N(0,1)$ is shared across patient $i$'s visits, and $\epsilon_{hij} \sim N(0,1)$ is visit-level noise. All components are independent.

Since the variance sums to 1, each $Z_{hij} \sim N(0,1)$ marginally. Setting $Y_{hij} = \mathbf{1}(Z_{hij} < \Phi^{-1}(p_{hij}))$ where $p_{hij} = \text{logit}^{-1}(\mathbf{x}_{hij}'\boldsymbol{\beta})$ guarantees that $P(Y_{hij}=1) = p_{hij}$ exactly.

The correlation structure is:

- **Same patient, different visits:** $\text{Cor}(Z_{hij}, Z_{hij'}) = \rho_P$
- **Different patients, same hospital:** $\text{Cor}(Z_{hij}, Z_{hi'j'}) = \rho_H$
- **Different hospitals:** $0$

The constraint $\rho_P \geq \rho_H \geq 0$ must hold.

```{r}
#| label: data-generation
#| code-fold: true
#| code-summary: "Data-generating function"

generate_data <- function(rho_H, rho_P, n_hospitals = 15,
                          n_patients_per_hosp = 15,
                          beta = c(-1, 1),
                          include_teaching = FALSE,
                          beta_teach = 0.5) {
  
  stopifnot(rho_P >= rho_H, rho_H >= 0, rho_P <= 1)
  
  # Hospital-level latent draw
  U <- rnorm(n_hospitals)
  
  # Hospital-level covariate (half are teaching hospitals)
  teaching <- rep(0L, n_hospitals)
  if (include_teaching) {
    teaching[sample(1:n_hospitals, floor(n_hospitals / 2))] <- 1L
  }
  
  patients <- data.frame(
    hosp_id = rep(1:n_hospitals, each = n_patients_per_hosp),
    pat_id  = 1:(n_hospitals * n_patients_per_hosp)
  )
  
  visits_list <- lapply(patients$pat_id, function(p) {
    h     <- patients$hosp_id[patients$pat_id == p]
    n_vis <- sample(2:6, 1)
    x     <- rnorm(n_vis)
    
    # Marginal linear predictor
    eta <- beta[1] + beta[2] * x
    if (include_teaching) eta <- eta + beta_teach * teaching[h]
    prob_marg <- plogis(eta)
    
    # Nested latent normal
    W   <- rnorm(1)
    eps <- rnorm(n_vis)
    Z   <- sqrt(rho_H) * U[h] +
           sqrt(rho_P - rho_H) * W +
           sqrt(1 - rho_P) * eps
    
    y <- as.integer(Z < qnorm(prob_marg))
    
    data.frame(hosp_id  = h,
               pat_id   = paste0(h, "_", p),
               teaching = teaching[h],
               x = x, y = y)
  })
  
  bind_rows(visits_list) |> arrange(hosp_id, pat_id)
}
```

# Federated GEE Implementation

## Protocol Overview

The federated GEE uses independence working correlation with a sandwich variance estimator clustered at the hospital (site) level. The protocol is iterative Fisher scoring:

1. **Coordinating center** broadcasts current $\hat{\boldsymbol{\beta}}$ to all sites.
2. **Each site** $k$ computes and transmits:
    - Bread: $B_k = \sum_{i \in k} \mathbf{X}_i' \text{diag}(\hat{\mu}_i(1-\hat{\mu}_i)) \, \mathbf{X}_i$
    - Score: $\mathbf{S}_k = \sum_{i \in k} \mathbf{X}_i'(\mathbf{Y}_i - \hat{\boldsymbol{\mu}}_i)$
    - Meat:  $M_k = \mathbf{S}_k \mathbf{S}_k'$
3. **Coordinating center** aggregates: $B = \sum_k B_k$, $\mathbf{S} = \sum_k \mathbf{S}_k$, $M = \sum_k M_k$.
4. **Update:** $\hat{\boldsymbol{\beta}}^{\text{new}} = \hat{\boldsymbol{\beta}}^{\text{old}} + B^{-1} \mathbf{S}$.
5. Repeat until convergence.
6. **Sandwich variance:** $\widehat{\text{Var}}(\hat{\boldsymbol{\beta}}) = B^{-1} M B^{-1}$.

No patient-level data ever leaves a site. Only the $p \times p$ matrices $B_k$, $M_k$ and the $p \times 1$ vector $\mathbf{S}_k$ are transmitted.

## Implementation

```{r}
#| label: federated-gee
#| code-fold: true
#| code-summary: "Federated GEE functions"

# ------------------------------------------------------------------
# Site-level computation: what each hospital computes and transmits
# ------------------------------------------------------------------
site_contribute <- function(site_data, beta, x_vars) {
  X  <- cbind(Intercept = 1, as.matrix(site_data[, x_vars, drop = FALSE]))
  y  <- site_data$y
  
  eta   <- as.numeric(X %*% beta)
  mu    <- plogis(eta)
  w     <- mu * (1 - mu)
  resid <- y - mu
  
  # Bread contribution: X'WX
  B_k <- crossprod(X * w, X)
  
  # Score contribution: X'(Y - mu)
  S_k <- crossprod(X, resid)
  
  # Meat contribution: S_k S_k' (cluster-level outer product)
  M_k <- tcrossprod(S_k)
  
  list(B = B_k, S = S_k, M = M_k)
}

# ------------------------------------------------------------------
# Coordinating center: aggregates site contributions and updates
# ------------------------------------------------------------------
federated_gee <- function(data, x_vars, max_iter = 50, tol = 1e-10) {
  
  sites <- split(data, data$hosp_id)
  p     <- 1 + length(x_vars)  # intercept + covariates
  beta  <- rep(0, p)
  names(beta) <- c("(Intercept)", x_vars)
  
  # Iterative Fisher scoring
  for (iter in 1:max_iter) {
    contribs <- lapply(sites, site_contribute, beta = beta, x_vars = x_vars)
    
    B_total <- Reduce("+", lapply(contribs, `[[`, "B"))
    S_total <- Reduce("+", lapply(contribs, `[[`, "S"))
    
    delta <- solve(B_total, S_total)
    beta  <- beta + as.numeric(delta)
    
    if (max(abs(delta)) < tol) break
  }
  
  # Final sandwich computation at converged beta
  contribs <- lapply(sites, site_contribute, beta = beta, x_vars = x_vars)
  B_total  <- Reduce("+", lapply(contribs, `[[`, "B"))
  M_total  <- Reduce("+", lapply(contribs, `[[`, "M"))
  
  B_inv <- solve(B_total)
  V     <- B_inv %*% M_total %*% B_inv
  
  list(
    coefficients = beta,
    se           = sqrt(diag(V)),
    vcov         = V,
    bread        = B_total,
    meat         = M_total,
    iterations   = iter
  )
}
```

# Simulation 1: Federated GEE Equals Pooled GEE

The first simulation establishes that the federated protocol produces results **numerically identical** to fitting GEE on the pooled data. This is not an approximation---the estimating equations decompose exactly.

## Single-Dataset Demonstration

```{r}
#| label: sim1-single
#| code-fold: true
#| code-summary: "Exact equivalence on one dataset"

d <- generate_data(rho_H = 0.2, rho_P = 0.5, n_hospitals = 20)

# Federated
fed_fit <- federated_gee(d, x_vars = "x")

# Pooled (geeglm with independence working correlation, hospital clustering)
pooled_fit <- geeglm(y ~ x, id = hosp_id, data = d,
                     family = binomial, corstr = "independence")

# Compare coefficients
comparison <- data.frame(
  Parameter       = names(fed_fit$coefficients),
  Federated       = fed_fit$coefficients,
  Pooled          = coef(pooled_fit),
  Difference      = fed_fit$coefficients - coef(pooled_fit),
  Fed_SE          = fed_fit$se,
  Pooled_SE       = sqrt(diag(vcov(pooled_fit))),
  SE_Difference   = fed_fit$se - sqrt(diag(vcov(pooled_fit)))
)
rownames(comparison) <- NULL
knitr::kable(comparison, digits = 12,
             caption = "Federated vs. pooled GEE: single dataset comparison")
```

## Across Parameter Settings

```{r}
#| label: sim1-sweep
#| code-fold: true
#| code-summary: "Equivalence across correlation settings"

settings <- expand.grid(rho_H = c(0, 0.1, 0.3), rho_P = c(0.2, 0.5, 0.8))
settings <- settings[settings$rho_P >= settings$rho_H, ]

equiv_results <- lapply(1:nrow(settings), function(i) {
  d <- generate_data(rho_H = settings$rho_H[i], rho_P = settings$rho_P[i],
                     n_hospitals = 10)
  
  fed  <- federated_gee(d, x_vars = "x")
  pool <- geeglm(y ~ x, id = hosp_id, data = d,
                 family = binomial, corstr = "independence")
  
  data.frame(
    rho_H         = settings$rho_H[i],
    rho_P         = settings$rho_P[i],
    max_beta_diff = max(abs(fed$coefficients - coef(pool))),
    max_se_diff   = max(abs(fed$se - sqrt(diag(vcov(pool)))))
  )
}) |> bind_rows()

knitr::kable(equiv_results, digits = 3,
             caption = "Maximum absolute difference in coefficients and SEs across settings",
             col.names = c("$\\rho_H$", "$\\rho_P$", 
                           "Max $|\\Delta\\hat{\\beta}|$",
                           "Max $|\\Delta \\text{SE}|$"),
             escape = FALSE)
```

The differences should be at machine precision ($\sim 10^{-12}$ or smaller). This is exact decomposition, not approximation.

# Simulation 2: GEE vs GLMM Under Nested Clustering

This simulation uses the nested copula DGP where the true **marginal** $\beta_1 = 1$. We fix $\rho_P = 0.5$ and sweep $\rho_H$ from 0 to 0.45 to show how increasing hospital-level heterogeneity affects the GLMM vs. GEE divergence.

```{r}
#| label: sim2-run
#| code-fold: true
#| code-summary: "Monte Carlo: GEE vs GLMM across hospital correlation"

n_sims <- 30  # Increase for smoother bands

rho_H_seq <- seq(0, 0.45, by = 0.09)
rho_P_fixed <- 0.5

res_sim2 <- list()

for (rH in rho_H_seq) {
  for (i in 1:n_sims) {
    d <- generate_data(rho_H = rH, rho_P = rho_P_fixed, n_hospitals = 20)
    
    # Federated GEE
    fed <- federated_gee(d, x_vars = "x")
    beta_gee <- fed$coefficients["x"]
    
    # GLMM
    beta_glmm <- tryCatch({
      mod <- glmer(y ~ x + (1 | hosp_id) + (1 | pat_id),
                   data = d, family = binomial, nAGQ = 1)
      fixef(mod)["x"]
    }, error = function(e) NA)
    
    res_sim2[[length(res_sim2) + 1]] <- data.frame(
      rho_H = rH, sim = i, GEE = beta_gee, GLMM = beta_glmm
    )
  }
}

df_sim2 <- bind_rows(res_sim2) |> drop_na() |>
  pivot_longer(c(GEE, GLMM), names_to = "Model", values_to = "Est") |>
  group_by(rho_H, Model) |>
  summarise(mean = mean(Est),
            lwr  = quantile(Est, 0.025),
            upr  = quantile(Est, 0.975), .groups = "drop")
```

```{r}
#| label: fig-sim2
#| fig-cap: "GEE recovers the true marginal β = 1 at all levels of hospital correlation. GLMM inflates as between-hospital heterogeneity grows."
#| fig-width: 7
#| fig-height: 4.5

ggplot(df_sim2, aes(x = rho_H, y = mean, color = Model, fill = Model)) +
  geom_line(linewidth = 1) +
  geom_ribbon(aes(ymin = lwr, ymax = upr), alpha = 0.2, color = NA) +
  geom_hline(yintercept = 1, linetype = "dashed") +
  labs(title = "Nested Copula DGP (True Marginal β = 1, ρ_P = 0.5)",
       x = bquote("Hospital-Level Correlation (" * rho[H] * ")"),
       y = "Coefficient Estimate") +
  theme_bw() +
  theme(legend.position = "bottom")
```

The GEE line stays flat at 1 because the marginal model is correct by construction. The GLMM is estimating the **conditional** effect, which must be larger than the marginal effect for logistic models---and the gap grows with $\rho_H$ because greater hospital-level heterogeneity means more attenuation when marginalizing.


# Simulation 3: Hospital-Level Covariates in Federated GEE

A hospital-level covariate (e.g., `teaching` status) is constant within each site. This makes the local design matrix rank-deficient---the `teaching` column is collinear with the intercept at every site. We demonstrate that federated GEE still recovers the coefficient because $B_k$ is summed before inversion, not inverted locally.

The true DGP uses $\beta_0 = -1$, $\beta_x = 1$, $\beta_\text{teach} = 0.5$.

```{r}
#| label: sim3-run
#| code-fold: true
#| code-summary: "Monte Carlo: recovery of hospital-level covariate"

n_sims <- 30

res_sim3 <- list()

for (i in 1:n_sims) {
  d <- generate_data(rho_H = 0.2, rho_P = 0.5, n_hospitals = 30,
                     include_teaching = TRUE, beta_teach = 0.5)
  
  # Federated GEE with both x and teaching
  fed <- federated_gee(d, x_vars = c("x", "teaching"))
  
  # Pooled GEE for verification
  pool <- geeglm(y ~ x + teaching, id = hosp_id, data = d,
                  family = binomial, corstr = "independence")
  
  res_sim3[[i]] <- data.frame(
    sim = i,
    fed_x        = fed$coefficients["x"],
    fed_teach    = fed$coefficients["teaching"],
    fed_se_teach = fed$se[3],
    pool_x       = coef(pool)["x"],
    pool_teach   = coef(pool)["teaching"],
    pool_se_teach = sqrt(vcov(pool)["teaching", "teaching"]),
    # Check local rank deficiency
    n_teaching_hospitals = sum(tapply(d$teaching, d$hosp_id, unique))
  )
}

df_sim3 <- bind_rows(res_sim3)
```

```{r}
#| label: fig-sim3
#| fig-cap: "Left: Federated GEE recovers the true teaching effect (0.5) despite within-site collinearity. Right: Federated and pooled estimates are identical."
#| fig-width: 10
#| fig-height: 4.5

p3a <- ggplot(df_sim3, aes(x = fed_teach)) +
  geom_histogram(bins = 15, fill = "steelblue", alpha = 0.7, color = "white") +
  geom_vline(xintercept = 0.5, linetype = "dashed", linewidth = 1) +
  labs(title = "Distribution of Teaching Coefficient",
       subtitle = "Federated GEE, true value = 0.5",
       x = bquote(hat(beta)[teach]), y = "Count") +
  theme_bw()

p3b <- ggplot(df_sim3, aes(x = pool_teach, y = fed_teach)) +
  geom_point(alpha = 0.7) +
  geom_abline(slope = 1, intercept = 0, linetype = "dashed") +
  labs(title = "Federated vs. Pooled",
       x = "Pooled GEE Estimate", y = "Federated GEE Estimate") +
  theme_bw()

p3a + p3b
```

```{r}
#| label: sim3-summary

cat("Teaching coefficient across", n_sims, "simulations:\n")
cat("  Mean estimate:", round(mean(df_sim3$fed_teach), 3), "\n")
cat("  True value:    0.500\n")
cat("  Mean SE:      ", round(mean(df_sim3$fed_se_teach), 3), "\n")
cat("  Empirical SD: ", round(sd(df_sim3$fed_teach), 3), "\n")
cat("  Max |fed - pooled| difference:",
    formatC(max(abs(df_sim3$fed_teach - df_sim3$pool_teach)), format = "e", digits = 3), "\n")
```

The key point: every site's $B_k$ is singular (teaching column is proportional to the intercept column locally), but the **sum** $B = \sum_k B_k$ is full rank because different sites have different values of `teaching`. The same logic applies to any hospital-level covariate: bed count, region, academic affiliation.


# Simulation 4: Sandwich Estimator Performance with Few Clusters

The sandwich variance estimator is consistent as the number of independent clusters grows. With few hospitals, it can underestimate variance, leading to undercoverage of confidence intervals. This simulation varies the number of hospitals and checks empirical coverage of nominal 95% Wald intervals.

```{r}
#| label: sim4-run
#| code-fold: true
#| code-summary: "Monte Carlo: coverage vs. number of hospitals"

n_sims <- 200  # Need more sims for stable coverage estimates

n_hosp_seq <- c(5, 10, 15, 30, 50, 100)

res_sim4 <- list()

for (K in n_hosp_seq) {
  for (i in 1:n_sims) {
    d <- generate_data(rho_H = 0.2, rho_P = 0.5, n_hospitals = K)
    
    fed <- tryCatch({
      federated_gee(d, x_vars = "x")
    }, error = function(e) NULL)
    
    if (is.null(fed)) next
    
    beta_hat <- fed$coefficients["x"]
    se_hat   <- fed$se[2]  # SE for x
    
    # 95% Wald CI
    ci_lwr <- beta_hat - 1.96 * se_hat
    ci_upr <- beta_hat + 1.96 * se_hat
    covers <- (ci_lwr <= 1) & (1 <= ci_upr)
    
    res_sim4[[length(res_sim4) + 1]] <- data.frame(
      n_hospitals = K, sim = i,
      beta_hat = beta_hat, se_hat = se_hat,
      covers = covers
    )
  }
}

df_sim4 <- bind_rows(res_sim4) |>
  group_by(n_hospitals) |>
  summarise(
    coverage     = mean(covers),
    mean_se      = mean(se_hat),
    empirical_sd = sd(beta_hat),
    se_ratio     = mean(se_hat) / sd(beta_hat),
    n_converged  = n(),
    .groups = "drop"
  )
```

```{r}
#| label: fig-sim4
#| fig-cap: "Empirical coverage of 95% Wald intervals based on the sandwich SE. With few hospitals, the sandwich underestimates variance and coverage drops below nominal."
#| fig-width: 7
#| fig-height: 4.5

ggplot(df_sim4, aes(x = n_hospitals, y = coverage)) +
  geom_line(linewidth = 1, color = "steelblue") +
  geom_point(size = 3, color = "steelblue") +
  geom_hline(yintercept = 0.95, linetype = "dashed") +
  geom_hline(yintercept = 0.95 - 1.96 * sqrt(0.95 * 0.05 / n_sims),
             linetype = "dotted", color = "grey50") +
  geom_hline(yintercept = 0.95 + 1.96 * sqrt(0.95 * 0.05 / n_sims),
             linetype = "dotted", color = "grey50") +
  scale_y_continuous(limits = c(0.75, 1), labels = scales::percent) +
  labs(title = "Sandwich Estimator Coverage vs. Number of Hospitals",
       subtitle = bquote("True marginal " * beta[x] * " = 1, " *
                         rho[H] * " = 0.2, " * rho[P] * " = 0.5"),
       x = "Number of Hospitals (Clusters)",
       y = "Empirical Coverage of 95% CI") +
  theme_bw()
```

```{r}
#| label: sim4-table

knitr::kable(df_sim4, digits = 3,
             col.names = c("Hospitals", "Coverage", "Mean SE",
                           "Empirical SD", "SE/SD Ratio", "N Converged"),
             caption = "Sandwich estimator performance by number of clusters. SE/SD Ratio < 1 indicates the sandwich underestimates true variability.")
```

## Interpretation

When the SE/SD ratio is below 1, the sandwich estimator is systematically too optimistic, and coverage falls below 95%. This matters for the VA application: with $\sim$130--170 medical centers, the standard sandwich should perform well. But if a subset analysis restricts to, say, 10--15 hospitals in a particular region, bias-corrected sandwich estimators (e.g., Mancl--DeRouen or Fay--Graubard corrections) should be used.

The correction multiplies the residual for cluster $k$ by $(I - H_{kk})^{-1/2}$ where $H_{kk}$ is the leverage of cluster $k$, analogous to the HC2/HC3 corrections in heteroskedasticity-robust inference for OLS. These corrections can be computed within the same federated protocol---each site can compute its own leverage contribution and apply the correction locally before transmitting the meat matrix.

# Summary of Simulations

| Simulation | Question | Key Finding |
|------------|----------|-------------|
| **1: Exact Equivalence** | Does federated GEE match pooled GEE? | Yes---numerically identical. This is exact decomposition, not approximation. |
| **2: GEE vs GLMM** | How do coefficients diverge with hospital-level clustering? | GEE recovers the marginal truth; GLMM inflates. Gap grows with $\rho_H$. |
| **3: Hospital Covariates** | Can federated GEE estimate site-constant predictors? | Yes---local singularity is resolved by summing $B_k$ before inversion. |
| **4: Sandwich Coverage** | How many hospitals are needed for reliable inference? | Coverage approaches nominal with 30+ clusters. Below 15, use bias-corrected sandwich. |

: Summary of simulation findings
